Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict} where
we can branch to. In particular, in Section \ref{ssec:ssa-norm}, we will
formalize interconversion with traditional SSA and show how it is semantically
sound to simply merge all nested scopes. The point of the slightly modified
syntax is simply to make some of our metatheoretic results more practical to
state.

@@

\subsection{Syntactic Metatheory}

@@

\TODO{some text about rewriting?}

@@

\subsection{Normalization}

\label{ssec:ssa-norm}

@@

@@

\TODO{some text about terms with holes}

@@

\TODO{some text about term with hole typing rules}

@@

\TODO{some text about block with hole typing rules}

@@

\begin{theorem}[Rewriting]
  Given \(\isrw{\mc{H}}{H}{I}\), we have
  \begin{itemize}
    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
  \end{itemize}

  Furthermore, we have that 
  \begin{itemize}
    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
  \end{itemize}
\end{theorem}

>>>>

For our type theoretic presentation of SSA, we will treat a very slight
generalization of the standard basic blocks with arguments syntax, in that we
will support \textit{nested scopes} and, in particular, \textit{if-statements}.
The program in Figure \ref{fig:bb-arg} translated to our syntax is given in
Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an
if-statement can be viewed as syntactic sugar for a conditional branch to one of
two anonymous basic blocks, while nested scopes simply \textit{restrict} where
we can branch to. 
% In particular, in Section \ref{ssec:ssa-norm}, we will formalize
% interconversion with traditional SSA and show how it is semantically sound to
% simply merge all nested scopes. The point of the slightly modified syntax is
% simply to make some of our metatheoretic results more practical to state.

@@

\subsection{Weakening and Substitution}

@@

\TODO{small rewriting section here?}

@@

% \subsection{Normalization}

% \label{ssec:ssa-norm}

% \TODO{this}

@@

\subsection{Syntax}

To properly be able to reason about congruence and congruence-based
optimization, we need to introduce the concept of \textit{terms with holes},
which are defined using the grammar in Figure
\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
syntactic categories with \textit{holes} of the form \(\lhole{X}\),
\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
introduce the concept of a \textit{hole-context} \(H\), which types a hole as being
one of:
\begin{itemize}
  \item \todo{expression}
  \item \todo{block}
  \item \todo{label-set}
\end{itemize}
\TODO{hole typing rules for expressions, blocks, label-sets; everything else is
the same}.

@@

@@

@@

@@

\begin{theorem}[Rewriting]
  Given \(\isrw{\mc{H}}{H}{I}\), we have
  \begin{itemize}
    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
  \end{itemize}
  Furthermore, we have that 
  \begin{itemize}
    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
  \end{itemize}
\end{theorem}

\subsection{Semantics}
